---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = c("svglite", "cairo_pdf"),
  out.width = "100%",
  dpi = 300,
  fig.retina = 3
)
```

# affectOU: Ornstein-Uhlenbeck model for Affect Dynamics

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/KCEvers/affectOU/graph/badge.svg)](https://app.codecov.io/gh/KCEvers/affectOU)
[![R-CMD-check](https://github.com/KCEvers/affectOU/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/KCEvers/affectOU/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The aim of `affectOU` is to provide tools for using the Ornstein-Uhlenbeck (OU) process in R with a chief focus on capturing affect dynamics, or how feelings change over time. The OU process is widely used for this purpose (e.g., Guthier et al., 2020; Kuppens et al., 2010a; Voelkle & Oud, 2013), formulating three core psychological mechanisms governed by three main sets of parameters (Uhlenbeck & Ornstein, 1930; Oravecz et al., 2011):

- **Drift matrix $\Theta$:** Governs the rate at which affect returns to baseline, therefore capturing a person's _emotion regulation capacity_ or _emotional inertia_;
- **Attractor $\mu$:** Represents the long-term average level of affect that the system tends to return to, therefore capturing a person's _baseline mood_ or _emotional setpoint_;
- **Diffusion matrix $\Gamma$:** Controls the magnitude of short-term fluctuations or noise in the affective system, representing an individual's sensitivity to environmental perturbations.

Rather than being a full-fledged package itself, `affectOU` rather serves as a demonstration of how on can package computational models, as detailed in Evers & Vanhasbroeck (2026).



## Installation

`affectOU` can be installed from [GitHub](https://github.com/KCEvers/affectOU) with:

``` r
# install.packages("pak")
pak::pak("KCEvers/affectOU")
```

Once installed, one can use the package through calling the `library()` function.

```{r library}
library(affectOU)
```

## Quick Start

The chief functionalities of the package revolve around a user-specified OU model of a particular dimensionality. To create such a model, one should use the [`affectOU()`](https://kcevers.github.io/affectOU/reference/affectOU.html) function and specify its dimensionality and/or parameters (see [Get Started](https://kcevers.github.io/affectOU/articles/affectOU.html)). For example, the following call creates a two-dimensional OU model with default parameter values:
```{r example}
# Create two-dimensional OU
model <- affectOU(ndim = 2)
model
```

To simulate data from this model, one should use the [`simulate()`]((https://kcevers.github.io/affectOU/reference/simulate.affectOU.html)) function:

```{r seed, echo=FALSE}
# Set seed for reproducibility
set.seed(
  123, 
  kind = "Mersenne-Twister", 
  normal.kind = "Inversion", 
  sample.kind = "Rejection"
)
```

```{r sim, echo = TRUE, results='hide', fig.show='hide', fig.height=5, fig.width=11}
# Simulate timeseries
sim <- simulate(model)
```

To visualize this simulation, one uses the [`plot()`]() function specifying the type of plot one wishes to create. For example, for a time-series plot, one specifies `type = "time"`:

```{r plot}
#| fig.height = 6,
#| fig.width = 11,
#| fig.alt = "Visualization of a time-series plot, showing the results of the simulation. In one time-series plot, it shows the simulated values for both dimensions of the OU process together with a horizontal line denoting the baseline affective state $\\mu$."
plot(sim, type = "time")
```



## Explore the OU Process

For more detailed examples and visual demonstrations of the model's characteristics, see the vignettes:

* [Get Started](https://kcevers.github.io/affectOU/articles/affectOU.html)
* [Affect Characteristics Implied by the OU Process](https://kcevers.github.io/affectOU/articles/characteristics.html)

Theoretical properties of the OU process are furthermore explained in the documentation for the functions [`stability()`](https://kcevers.github.io/affectOU/reference/stability.affectOU.html), [`stationary()`](https://kcevers.github.io/affectOU/reference/stationary.affectOU.html), and [`relaxation()`](https://kcevers.github.io/affectOU/reference/relaxation.affectOU.html). For fitting unidimensional OU models to data, see [`fit()`](https://kcevers.github.io/affectOU/reference/fit.affectOU.html).



## References

Evers, K. C. & Vanhasbroeck, N. (2026). _Sharing computational models as reproducible and user-friendly packages: A tutorial in R._ Manuscript in preparation.

Guthier, C., Dormann, C., & Voelkle, M. C. (2020). Reciprocal effects between job stressors and burnout: A continuous-time meta-analysis of longitudinal studies. _Psychological Bulletin, 146_, 1146-1173. doi: [10.1037/bul0000304](https://doi.org/10.1037/bul0000304)

Kuppens, P., Oravecz, Z., & Tuerlinckx, F. (2010). Feelings change: Accounting for individual differences in the temporal dynamics of affect. _Journal of Personality and Social Psychology, 99_, 1042-1060. doi: [10.1037/a0020962](https://doi.org/10.1037/a0020962)

Oravecz, Z., Tuerlinckx, F., & Vandekerckhove, J. (2011). A hierarchical latent stochastic differential equation model for affective dynamics. _Psychological Methods, 16_, 468-490. doi: [10.1037/a0024375](https://doi.org/10.1037/a0024375)

Uhlenbeck, G. E. & Ornstein, L. S. (1930). On the theory of Brownian motion. _Physical Review, 46_, 823-841.

Voelkle, M. C. & Oud, J. H. L. (2013). Continuous time modelling with individually varying time intervals for oscillating and non-oscillating processes. _British Journal of Mathematical and Statistical Psychology, 66_, 103-126. doi: [10.1111/j.2044-8317.2012.02043.x](https://doi.org/10.1111/j.2044-8317.2012.02043.x)

