---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = c("svglite", "cairo_pdf"),
  out.width = "100%",
  dpi = 300,
  fig.retina = 3
)
```

# affectOU: Ornstein-Uhlenbeck Model for Affect Dynamics

<!-- badges: start -->
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.20037783.svg)](https://doi.org/10.5281/zenodo.20037783)
[![Codecov test coverage](https://codecov.io/gh/kcevers/affectOU/graph/badge.svg)](https://app.codecov.io/gh/kcevers/affectOU)
[![R-CMD-check](https://github.com/kcevers/affectOU/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/kcevers/affectOU/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The aim of `affectOU` is to provide tools for simulating the Ornstein-Uhlenbeck (OU) process in R. The OU is widely used to model affect dynamics: how feelings or emotions change over time (e.g., Guthier et al., 2020; Kuppens et al., 2010a; Voelkle & Oud, 2013). It formulates three core psychological mechanisms through three sets of parameters (Uhlenbeck & Ornstein, 1930; Oravecz et al., 2011):

- **Drift matrix Θ:** Governs the rate at which affect returns to baseline, capturing a person's _emotion regulation capacity_ or _emotional inertia_;
- **Attractor μ:** The long-term average level of affect that the system tends to return to, representing a person's _baseline mood_ or _emotional setpoint_;
- **Diffusion matrix Γ:** Controls the magnitude of short-term fluctuations or noise in the affective system, characterizing an individual's _reactivity_ or _sensitivity_ to environmental perturbations.

`affectOU` primarily serves as a demonstration of a packaged computational model, as detailed in Evers & Vanhasbroeck (2026), “Sharing Computational Models as Reproducible and User-Friendly Packages: A Tutorial in R” (forthcoming).

## Installation

`affectOU` can be installed from [GitHub](https://github.com/kcevers/affectOU) with:

``` r
# install.packages("pak")
pak::pak("kcevers/affectOU")
```

Once installed, one can use the package through calling the `library()` function.

```{r library}
library(affectOU)
```

## Quick Start

<!-- The chief functionalities of the package revolve around a user-specified OU model of a particular dimensionality. To create such a model, one should use the [`affectOU()`](https://kcevers.github.io/affectOU/reference/affectOU.html) function and specify its dimensionality and/or parameters (see [Get Started](https://kcevers.github.io/affectOU/articles/affectOU.html)). For example, the following call creates a two-dimensional OU model with default parameter values: -->

The core functionality of the package revolves around an OU model object created by the function [`affectOU()`](https://kcevers.github.io/affectOU/reference/affectOU.html). For example, the following creates a two-dimensional model with default parameter values:

```{r example}
model <- affectOU(ndim = 2)
```

```{r example-print}
print(model)
```

```{r example-summary}
summary(model)
```

The model can then be simulated with [`simulate()`](https://kcevers.github.io/affectOU/reference/simulate.affectOU.html) and visualised over time with [`plot()`](https://kcevers.github.io/affectOU/reference/plot.simulate_affectOU.html).

```{r seed, echo=FALSE}
# Set seed for reproducibility
set.seed(
  123,
  kind = "Mersenne-Twister",
  normal.kind = "Inversion",
  sample.kind = "Rejection"
)
```

```{r sim, echo = TRUE}
sim <- simulate(model)
```

```{r tutorial_paper, echo = FALSE, include=FALSE, fig.show='hide', fig.height=4, fig.width=7}
plot(sim,
  main = "Ornstein-Uhlenbeck Affect Dynamics",
  mar = c(2, 2, 0.5, 2)
)
```

```{r proposal_paper, echo = FALSE, include=FALSE, fig.show='hide', fig.height=4, fig.width=6.25}
if (requireNamespace("extrafont")) {
  library(extrafont)

  # # Import fonts once
  # font_import()
  # # Look at available fonts
  # fonts()

  # Register fonts for PDF output
  loadfonts(device = "pdf", quiet = TRUE)

  plot(sim,
    main = "", mar = c(0, 0, 0, 0), xlim = c(0, 30), cex.lab = 1.3,
    #  xaxt = "n", yaxt = "n",
    family = "Georgia", col.grid = "white", col.axis = "white"
  )
}
```

```{r plot, results='hide', fig.show='hide'}
plot(sim)
```

```{r fig, echo = FALSE, include=TRUE, fig.height=4, fig.width=11}
#| fig.alt = "Visualization of a time-series plot, showing the results of the simulation. In one time-series plot, it shows the simulated values for both dimensions of the OU process together with a horizontal line denoting the baseline affective state $\\mu$."
plot(sim, mar = c(2, 2, 0.5, 2))
```


## Explore the OU Process

For more detailed examples and visual demonstrations of the model's characteristics, see the vignettes:

* [Get started with affectOU](https://kcevers.github.io/affectOU/articles/affectOU.html)
* [Affect characteristics implied by the OU process](https://kcevers.github.io/affectOU/articles/characteristics.html)

More ways to visualise the OU are offered in [`plot()`](https://kcevers.github.io/affectOU/reference/plot.simulate_affectOU.html). Its theoretical properties are further explained in the function documentation of [`stability()`](https://kcevers.github.io/affectOU/reference/stability.affectOU.html) and [`stationary()`](https://kcevers.github.io/affectOU/reference/stationary.affectOU.html). For fitting unidimensional OU models to data, see [`fit()`](https://kcevers.github.io/affectOU/reference/fit.affectOU.html).



## References

Evers, K. C. & Vanhasbroeck, N. (2026). _Sharing computational models as reproducible and user-friendly packages: A tutorial in R._ Manuscript in preparation.

Guthier, C., Dormann, C., & Voelkle, M. C. (2020). Reciprocal effects between job stressors and burnout: A continuous-time meta-analysis of longitudinal studies. _Psychological Bulletin, 146_, 1146-1173. doi: [10.1037/bul0000304](https://doi.org/10.1037/bul0000304)

Kuppens, P., Oravecz, Z., & Tuerlinckx, F. (2010). Feelings change: Accounting for individual differences in the temporal dynamics of affect. _Journal of Personality and Social Psychology, 99_, 1042-1060. doi: [10.1037/a0020962](https://doi.org/10.1037/a0020962)

Oravecz, Z., Tuerlinckx, F., & Vandekerckhove, J. (2011). A hierarchical latent stochastic differential equation model for affective dynamics. _Psychological Methods, 16_, 468-490. doi: [10.1037/a0024375](https://doi.org/10.1037/a0024375)

Uhlenbeck, G. E. & Ornstein, L. S. (1930). On the theory of Brownian motion. _Physical Review, 46_, 823-841.

Voelkle, M. C. & Oud, J. H. L. (2013). Continuous time modelling with individually varying time intervals for oscillating and non-oscillating processes. _British Journal of Mathematical and Statistical Psychology, 66_, 103-126. doi: [10.1111/j.2044-8317.2012.02043.x](https://doi.org/10.1111/j.2044-8317.2012.02043.x)

