---
title: "Get started with affectOU"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with affectOU}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

`affectOU` provides tools for modeling affect dynamics using the Ornstein-Uhlenbeck (OU) process -- a continuous-time stochastic model that captures how emotions fluctuate around a baseline and return to equilibrium over time.

This vignette walks through the package's core functionality: simulating the OU process, computing theoretical quantities, and fitting the unidimensional OU to data. See the vignette [Affect characteristics implied by the OU process](characteristics.html) for visual demonstrations of the model's qualitative behaviours and their psychological interpretations.


<!-- ## Setup -->

```{r setup, message = FALSE, warning = FALSE}
library(affectOU)
```

## Creating a Model


### One-Dimensional Model

For a univariate affective process, the stochastic differential equation of the OU is:

$$d{X}(t) = {\theta}({\mu} - {X}(t))dt + {\gamma} \, d{W}(t)$$


Here, the parameters represent:

* $\mu$ (`mu`): Attractor location, or the baseline and "set point" of affect;
* $\theta$ (`theta`): Attractor strength, or how quickly affect returns to the baseline; 
* $\gamma$ (`gamma`): Diffusion coefficient, or the scale of the random perturbations.

The diffusion coefficient $\gamma$ is more easily interpreted in terms of the noise variance $\sigma$ (`sigma`), where $\sigma = \gamma^2$. 

You can create a simple one-dimensional model with default parameters by calling the `affectOU` constructor:

```{r create-1d}
model <- affectOU()
model
```

Alternatively, you can specify custom parameters to represent, for example, a process with a stronger mean-reversion ($\theta = 1$), positive baseline ($\mu = 1$), and smaller amount of random perturbations ($\sigma = 0.25$):
 
```{r create-1d-custom}
model <- affectOU(theta = 1, mu = 1, sigma = 0.25)
model
```

### Multidimensional Models

The OU framework extends naturally to multiple affect dimensions (e.g., valence and arousal, positive and negative affect). For a d-dimensional process, the OU looks like:

$$d\mathbf{X}(t) = \mathbf{\Theta}(\mathbf{\mu} - \mathbf{X}(t))dt + \mathbf{\Gamma} \, d\mathbf{W}(t)$$

For multivariate models, the noise covariance matrix $\Sigma$ is defined as $\Sigma = \Gamma \Gamma^\top$.

By default, `affectOU()` creates a model with uncoupled dynamics (diagonal $\mathbf{\Theta}$ and $\mathbf{\Sigma}$) and identical parameters across dimensions. For example, for a two-dimensional model, we get:



```{r create-2d-default}
model_2d <- affectOU(ndim = 2)
model_2d
```

The [`update()`](../reference/update.affectOU.html) function can be used to modify an existing model without recreating it from scratch:

```{r create-2d}
model_2d <- update(
  model_2d,
  theta = diag(c(0.5, 0.3)), # Different regulation speeds
  mu = c(0, 1), # Different baselines
  sigma = diag(c(1, 0.5)) # Different noise levels
)

model_2d
```

To create coupled dynamics, in which dimensions affect each other over time, a non-diagonal $\mathbf{\Theta}$ matrix should be specified. For example, the following creates a model in which the first dimension has a stronger influence on the second than vice versa:

```{r create-2d-coupled}
theta_coupled <- c(
  0.5, 0.1,
  0.2, 0.3
) |>
  matrix(nrow = 2, ncol = 2, byrow = TRUE)

model_coupled <- update(
  model_2d,
  theta = theta_coupled
)

model_coupled
```

## Simulating Trajectories

You can simulate data from the specified model through the [`simulate()`](../reference/simulate.affectOU.html) function:

```{r simulate-basic}
sim <- simulate(model)
```

Though the OU is a continuous-time process, simulation requires evaluating it at discrete time points -- either via exact sampling from the known transition distribution or via numerical discretization methods such as Euler-Maruyama approximation, which is a common numerical approach for simulating stochastic differential equations. In the package, we use the latter. For the OU, it amounts to updating the state $\mathbf{X}$ after a discrete time-step $\Delta t$ according to the following equation:

\begin{equation}
\begin{split}
  \mathbf{X}_{t + \Delta t} &= \mathbf{X}_t + \mathbf{\Theta} (\mathbf{\mu} - \mathbf{X}_t) \Delta t + \mathbf{\Gamma} \mathbf{\epsilon}_t \sqrt{\Delta t} \\
  \mathbf{\epsilon}_t &\sim N(\mathbf{0}, \mathbf{I}_d)
\end{split}
\end{equation}

where $\mathbf{I}_d$ is the $d \times d$ identity matrix (i.e., a matrix with 1s on the diagonal and 0s elsewhere) and where $\Delta t$ is sufficiently small to ensure sensible results (e.g., $\Delta t = 0.01$).

Simulations can be customized to fit one's needs, specifically by changing the arguments:

* `nsim`: Number of replications or simulations to run; 
* `dt`: The (discretized) time step to use within the Euler-Maruyama approximation;
* `stop`: Total time to simulate for (e.g., `stop = 100` simulates from time 0 to time 100);
* `save_at`: Time interval between saved simulation values;
* `seed`: Seed to use for sampling random perturbations.

```{r simulate-advanced}
sim <- simulate(
  model,
  nsim = 1,
  dt = 0.01,
  stop = 100,
  save_at = 0.1,
  seed = 42
)

print(sim)
```


### Visualizing Simulations

A first way to explore the simulated data is by plotting it. `affectOU` provides the `plot` function with several different types of plots to visualize the simulated data, namely:

```{r plot-sim}
#| fig.show = "hold",
#| fig.height = 6,
#| out.width = "50%",
#| fig.alt = "Four plots are shown in a two by two fashion. In order, the plots contain (a) the time series of the simulated data, (b) the distribution of its values, (c) a visualization of the autocorrelations per lag in time, and (d) the phase portrait of how affect changed over time. All plots use the default visualization specifications, meaning the default titles explaining the content of the simulations ('Affect Dynamics', 'Affect Distribution', 'Autocorrelation Function', and 'Phase Portrait' respectively) and using the default colors of the package (using the 'Temps' palette of the package `grDevices`)."
# Time series trajectory (default)
plot(sim)

# Distribution of affect values
plot(sim, type = "histogram")

# Autocorrelation function with theoretical overlay
plot(sim, type = "acf")

# Phase portrait (lag-1 relationship)
plot(sim, type = "phase")
```

* Time series trajectory (`type = "time"`, [`ou_plot_time()`](../reference/ou_plot_time.html)): Shows the simulated values of affect over time, along with a horizontal line denoting the baseline affective state $\mu$.

* Distribution of affect values (`type = "histogram"`, [`ou_plot_histogram()`](../reference/ou_plot_histogram.html)): Shows the distribution of simulated affect values aggregated across time, along with a vertical line denoting the baseline affective state $\mu$.

* Autocorrelation function (`type = "acf"`, [`ou_plot_acf()`](../reference/ou_plot_acf.html)): Shows the value of the autocorrelation between simulated values at particular lags apart, along with a dashed line denoting the theoretical autocorrelation function implied by the OU model.

* Phase portrait (`type = "phase"`, [`ou_plot_phase()`](../reference/ou_plot_phase.html)): Shows how affect at time $t$ relates to the value of affect at the next time point $t + \Delta t$. The dashed line denotes the theoretical relationship between affect at time $t$ and $t + \Delta t$ implied by the OU model, which is a linear relationship with slope $e^{-\mathbf{\Theta} \Delta t}$ and intercept $(1 - e^{-\mathbf{\Theta} \Delta t} \mathbf{\mu})$.


### Multiple Dimensions 

The multidimensional OU can be simulated and visualised in the same way as before. Here, we set `by_dim = TRUE` to show each dimension in a separate subplot:

```{r simulate-md}
#| fig.height = 4,
#| fig.alt = "Time series plot with two panels, each showing the evolution of affect over time."
# Create an affectOU model with two dimensions
model_2D <- affectOU(ndim = 2)

# Simulate timeseries
sim <- simulate(
  model_2D
)

# Plot
plot(sim, by_dim = TRUE)
```



### Multiple Simulations

Multiple independent realizations of the model can be generated by setting `nsim` > 1. This showcases the variability in trajectories that arises from the stochastic nature of the OU process. For example, the following code generates three independent simulations from the same model:

```{r simulate-multiple}
#| fig.height = 5,
#| fig.alt = "Shown is a time-series plot containing all three of the simulations overlayed on top of each other."
sim <- simulate(model, nsim = 3)
plot(sim, type = "time")
```



### Extracting Simulation Data

Simulated data can be extracted in different formats:

```{r extract-data}
# Directly from the simulation object
data <- sim$data # 3D array (time × dimension × simulation)
```

```{r extract-data-head}
# Using head() or tail()
head(sim, n = 3)
```

```{r extract-data-list}
# As a list of data frames (one per simulation)
sim_list <- as.list(sim)
head(sim_list[[1]], n = 3)
```

```{r extract-data-df}
# As a data frame (long or wide format)
sim_df <- as.data.frame(sim)
head(sim_df, n = 3)
```

```{r extract-data-array}
# As a 3D array (time × dimension × simulation)
sim_array <- as.array(sim)
dim(sim_array)
```

```{r extract-data-matrix}
# As a matrix (time × dimension)
sim_matrix <- as.matrix(sim)
head(sim_matrix, n = 3)
```

## Theoretical Quantities

### Model Summary

Key theoretical properties can be computed with [`summary()`](../reference/summary.affectOU.html), showing the stability of the system (i.e., its long-term behavior) and the stationary distribution (i.e., long-run means and (co)variances) implied by the model parameters:

```{r summary}
summary(model)
```

For multidimensional models, the model summary also includes coupling indicators:

```{r summary-coupled}
summary(model_coupled)
```

More detail is provided by [`stability()`](../reference/stability.affectOU.html) and [`stationary()`](../reference/stationary.affectOU.html):

```{r stability}
stability(model_coupled)
```

```{r stationary}
stationary(model_coupled)
```


## Fitting Models to Data

[`fit()`](../reference/fit.affectOU.html) estimates OU parameters from univariate time series using maximum likelihood:
 
```{r fit}
# Generate data from known parameters
true_model <- affectOU(theta = 0.5, mu = 0, sigma = 1)
sim <- simulate(true_model, dt = 0.01, stop = 500, save_at = 0.1)

# Extract the simulated data
data <- as.data.frame(sim)

# Fit the model
fitted <- fit(true_model, data = data$value, times = data$time)
fitted
```


```{r plot-fit-residuals}
#| fig.alt = "Visualization of the residuals left in the data after fitting the OU to them as a typical diagonistic plot used for linear regression models."
# Residuals over time
plot(fitted, type = "residuals")
```

See [`plot()`](../reference/plot.fit_affectOU.html) for more details on available diagnostics and their interpretations.

## Extract Equation

To ease the process of writing down the OU's stochastic differential equation, `affectOU` includes an [`equation()`](../reference/equation.affectOU.html) function which renders the model's equation in multiple formats, offering plain text, LaTeX, R expression, or code output.

```{r equation-plain}
equation(model)

equation(model, type = "latex")
```

By default, this function will output the general equation and define the parameter values below. The parameter values can also directly be included in the equation by setting `inline = TRUE`:  

```{r equation-plain-inline}
equation(model, inline = TRUE)
```

## Summary

The typical workflow starts with the specification of an OU model of the [`affectOU()`](../reference/affectOU.html) class, optionally adjusting its parameters via its arguments or through the [`update()`](../reference/update.affectOU.html) method. Model trajectories can be generated with [`simulate()`](../reference/simulate.affectOU.html) and visualized with [`plot()`](../reference/plot.simulate_affectOU.html). The model may be further understood by inspecting its theoretical properties, for which the package offers [`summary()`](../reference/summary.affectOU.html), [`stability()`](../reference/stability.affectOU.html), and [`stationary()`](../reference/stationary.affectOU.html). Finally, the OU can be fitted to empirical data through the [`fit()`](../reference/fit.affectOU.html) method, and the resulting equation can be extracted with [`equation()`](../reference/equation.affectOU.html).

